---
title: "MATH 3080 Lab Project 8"
author: "Harrison Webb"
date: "03/30/2022"
output:
  html_document:
    toc: TRUE
---

# Problem 1 (Verzani problem 12.28)

*In the data set `mtcars` the variables `mpg`, `cyl`, and `am` indicate the
miles per gallon, the number of cylinders, and the type of transmission,
respectively. Perform a two-way ANOVA modeling `mpg` by the `cyl` and `am`, each
treated as a categorical variable.*

*Is there an indication of an interaction? Do both the number of cylinders and
the type of transmission make a difference?*

```{r, error = TRUE, fig.align='center', out.width='75%'}
with(mtcars, interaction.plot(cyl, am, mpg))
```
</br>
The interaction plot is not completely parallel, suggesting that an interaction may be happening, but there is no intersection, which would strongly suggest an interaction.

```{r}
model1 <- lm(mpg ~ as.factor(cyl) * as.factor(am), data = mtcars)
summary(model1)
```
The F-statistic p-value here is extremely small, so we can say with near-certainty that at least one treatment has an effect on the response variable (mpg). We see that the treatments with a statistically significant effect are having 8 cylinders (negative relationship with mpg) and having an automatic transmission (positive relationship with mpg).
</br>

# Problem 2 (Verzani problem 12.29)

*The data set `ToothGrowth` has measurements of tooth growth (`len`) of guinea
pigs for different dosages of Vitamin C (`dose`) and two different delivery
methods (`supp`).*

*Perform a two-way analysis of variance of tooth growth modeled by dosage and
delivery method. First, fit the full model including interactions and use the
$F$-test to compare this with the additive model.*

First lets make an interaction plot
```{r, error = TRUE, fig.align='center', out.width='75%'}
with(ToothGrowth, interaction.plot(dose, supp, len))
```
At doses 0.5 and 1 the plots appear parallel, but there is an intersection at dose = 2, which suggests at least some interaction.

```{r}
toothGrowthWithInteraction <- lm(len ~ supp * as.factor(dose), data = ToothGrowth)
toothGrowthNoInteraction <- lm(len ~ supp + as.factor(dose), data = ToothGrowth)

anova(toothGrowthNoInteraction, toothGrowthWithInteraction)
```
Our p value is significant at alpha = 0.05, so it appears there are interactions.
</br>

# Problem 3 (Verzani problem 12.30)

*The data set `OrchardSprays` contains measurements on the effectiveness of
various sprays on repelling honeybees. The variable `decrease` measures
effectiveness of the spray, `treatment` records the type of treatment, and
`rowpos` records the row in the field the measurement comes from.*

*Make an interaction plot of the mean of `decrease` with `treatment` as a trace
factor. Then fit the additive analysis-of-variance model and the model with
interaction. Compare the two models using `anova`. Is the interaction model
suggested by the results of the modeling?*

```{r, error = TRUE}
# Your code here
```

# Problem 4 (Verzani problem 12.31)

*What does R output when there is not enough data to estimate the parameters and
an error term? We check with a simple example. In checking a comment that
everything is better with butter on it, a student asks four people to rate the
four combinations of bread and corn with and without butter. The data collected
is:*

```{r}
d <- data.frame(
  rating = c(8, 6, 8, 4),
  food = gl(2, 2, 4, labels = c("bread", "corn")),
  butter = gl(2, 1, 4, labels = c("yes", "no"))
)
xtabs(rating ~ butter + food, d)
with(d, interaction.plot(butter, food, rating))
```

*Look at the summary of the multiplicative model*

```{r}
summary(lm(rating ~ butter * food, d))
```

*Do the coefficients get estimated? What is the issue then?*

**YOUR COMMENTS HERE**

