---
title: "R Notebook"
output: html_notebook
---

# Logistic Regression
Use when the response is Bernoulli (yes/no or true/false or 0/1)

$g(x) = \frac{e^x}{1+e^x}$

implement using function glm()
Very similar to lm()

Example: Predict probability that iris is versicolor, given it's dimensions
```{r}
fit <- glm(I(Species == "versicolor") ~ Sepal.Length + Sepal.Width + 
       Petal.Length + Petal.Width,
    family = binomial, data = iris) #use binomial for logistic models
summary(fit)
```
Differences in interpretation:
Increase in x gives $e^{\beta_j}$ increase in odds

Most functions from lm carry over:
summary, coef, predict

```{r}
exp(coef(fit))
```

# Nonlinear Regression
use nls() function

```{r}
#par(mfrow = c(1,2))
plot(conc~time, data = Indometh, pch = 20)
#plot(log(conc)~time, data = Indometh, pch = 20)
```
Must have some background idea of what the function looks like

Step 1: Define the function
```{r}
doubleExp <- function(c0, a1, a2, b1, b2, t){
   c0+a1*exp(-b1*t)+a2*exp(-b2*t)
}

#Step 2: Use nls
#step 2b: create initial guesses for your parameters
nlfit <- nls(conc~doubleExp(c0, a1, a2, b1, b2, time), data = Indometh, 
    start = c(c0=0.125, a1=0.75, a2=1, b1=1.5, b2=1.1))
summary(nlfit)

curve(doubleExp(x, c0 = coef(nlfit)["c0"],
                a1 = coef(nlfit)["a1"],
                b1 = coef(nlfit)["b1"],
                a2 = coef(nlfit)["a2"],
                b2 = coef(nlfit)["b2"]),
      from = 0, to = 8)
points(conc~time, data = Indometh)
```


