---
title: "Lab 5 Notes"
output: html_notebook
---

Week 5 Linear Models

## Review from last week
Prediction Intervals 
  - Interval estimating taking "one more sample" based on a parent sample
  -Dependent on sample distribution

Replication Interval - 

Tolerance Interval
   - Contains a certain percent of the data with a certain certainty

Nonparametric Interval
   - Dont assume anyting about distribution



## Linear Regression

Simple Linear Regression
$y_i = \beta_0 + \beta_1 * x + \epsilon_i$
Goal: estimate $\beta_0$ and $\beta_1$

Transformations can allow us to use linear models for many uses

### Basic Linear Model
```{r}
summary(iris)
model1 = lm(Sepal.Length~Sepal.Width, data = iris)
summary(model1)

#Plot
plot(Sepal.Length~Sepal.Width, data = iris, col = Species, pch = 20)
abline(model1)


```
### Fitting Different LInes for different species
```{r}
#Fit different lines per species
setosaModel = lm(Sepal.Length~Sepal.Width, data = iris, subset = Species == "setosa")
summary(setosaModel)
plot(Sepal.Length~Sepal.Width, data = iris, col = Species, pch = 20, subset = Species == "setosa")
abline(setosaModel)
```

### Suppressing the Intercept with a -1
```{r}
suppressedModel = lm(Sepal.Length~Sepal.Width-1, data = iris)
suppressedModel
#summary(suppressedModel)

```

lm() has some funky special operations. Use I() around any math expression to be interpreted literally
```{r}
lm(Sepal.Length~I(Sepal.Width^2), data = iris)
```


### Looking at the Model Summary
```{r}
setosaModel$coefficients #Get coefficients
setosaModel$residuals #get residuals
```

However, its better to access this info with the respective functions
```{r}
coef(setosaModel)
resid(setosaModel)
```


### Scale Function
Standardizes data: subtracts mean, divides by standard deviation
If forming a linear model on scaled data, interpretation is that a 1 sd increase in x leads to a 1 sd increase in y

```{r}
scaled = scale(iris$Sepal.Length)
mean(scaled)
sd(scaled)

lm(scale(Sepal.Length)~scale(Sepal.Width), data = iris)
```

### Predict Function
Allows for interpolation along the line
Arguments:
   1. lm result
   2. Must be a data frame
   3. Data frame name must match the originally used name

Don't abuse this for extrapolation

```{r}
predict(setosaModel, data.frame(Sepal.Width = c(2.7, 3.2, 3.7)))
```

