---
title: "MATH 3080 Lab Project 3"
author: "Harrison Webb"
date: "12/29/2019"
output:
  html_document:
    toc: TRUE
---

# Problem 1 (Verzani problem 12.2)

*For the data set `Cars93` (**MASS**) perform a one-way analysis of variance of
`MPG.highway` for each level of `DriveTrain`. Does the data support the null
hypothesis of equal population means? (Use `aov()` for this problem.)*

```{r, include=FALSE}
library(MASS)
```

#### Solution
```{r, error = TRUE, fig.cap="Variances appear to differ a bit more than we would like, but we will still carryout the test", fig.align='center'}
#check equal variances
boxplot(MPG.highway~DriveTrain, data = Cars93, col = c("darkgreen", "blue", "orange"))

#check normality
shapiro.test(Cars93$MPG.highway) #Shapiro test results do not strongly indicate a normal distribution, but we'll proceed anyway

#conduct ANOVA test
result = aov(MPG.highway~DriveTrain, data = Cars93)
summary(result)
```
We get a p-value of 0.00281 which is smaller than 0.05, so we reject $H_0$ and have reason to believe that the means of each group are not all the same.

</br>

# Problem 2 (Verzani problem 12.4)

*The data set `carsafety` (**UsingR**) contains car-crash data. For several
makes of cars the number of drivers killed per million is recorded in
`Drivers.deaths`. The number of drivers of other cars killed in accidents with
these cars, per million, is recorded in `Other.deaths`. The variable `type` is a
factor indicating the type of car.*

*Perform a one-way analysis of variance of the model `Drivers.deaths ~ type`. Is
there a difference in population means? Did you assume equal variances? Normally
distributed populations?*

*Repeat with an analysis of variance of the model `Other.deaths ~ type`. Is
there a difference in population means? (Use `oneway.test()` for this problem.)*
```{r, include=FALSE}
library(UsingR)
```
#### Solution
```{r, error = TRUE, out.width='85%', fig.align='center'}

#First we will check for normality of the data
shapiro.test(carsafety$Driver.deaths)

#Now construct a boxplot to look at variance of each type
par(mar=c(7,4,4,1))
boxplot(Driver.deaths~type, data = carsafety, las = 2, xlab = "", ylab = "Deaths", main = "Driver deaths vs Car Type", col = carsafety$type)
mtext("Car Type", 1, line = 5)

```
</br>
It appears that the groups of SUV, Compact, Large, and Midsize have similar variances, while Minivan, pickup, and Subcompact appear to have different variances. Additionally, the Shapiro test does not strongly indicate a normal distribution, but we'll proceed anyway.

With ANOVA, our null hypothesis $H_0$ is $\mu_1 = \mu_2 =  ... = \mu_n$, and the alternate hypothesis $H_a$ is that not all means are equal. Now, run the ANOVA test using oneway.test:
```{r}
oneway.test(Driver.deaths~type, data = carsafety, var.equal = TRUE)
```
We get a p-value of 0.01407 which is smaller than 0.05, so we reject $H_0$ and have reason to believe that the means of each group are not all the same.

# Problem 3 (Verzani problem 12.7)

*A manufacturer of point-of-sale merchandise tests three types of ENTER-button
markings. They wish to minimize wear, as customers get annoyed when the markings
on this button wear off. They construct a test of the three types, and conduct
several trials for each. The results, in unspecified units, are recorded in the
following table:*

|        |     |     |     |     |     |     |
|--------|-----|-----|-----|-----|-----|-----|
| Type 1 | 303 | 293 | 296 | 299 | 298 |     |
| Type 2 | 322 | 326 | 315 | 318 | 320 | 320 |
| Type 3 | 309 | 327 | 317 | 315 |     |     |

*Is there a difference in wear time among the three types? Answer this using a
one-way ANOVA.*

#### Solution

First we will create a data frame to make working with this data easier. Constructing a boxplot for each type, we see that the variances are pretty comparable between the three groups.
```{r, error = TRUE, fig.align='center'}
# Your code here
#make one vector for types and another for #s
df = data.frame(Type1 = c(303, 293, 296, 299, 298, NA),
                              Type2 = c(322, 326, 315, 318, 320, 320),
                              Type3 = c(309, 327, 317, 315, NA, NA))

df2= data.frame(type = c(1,1,1,1,1,2,2,2,2,2,2,3,3,3,3),
                value = c(303,293,296,299,298,322,326,315,318,320,320,309,327,317,315))
boxplot(value~type, data = df2, col = c("#4c23ba", "#d70081", "#ff3838"))
```
Next, check the normality of the data using Shapiro test:
```{r}
shapiro.test(df2$value)
```
We get a p-value of 0.1518, so let's assume that the data is approximately normal.
We have 'confirmed' our assumptions of equivariance and normality, and can now conduct the test.
```{r}
oneway.test(value~type, data = df2, var.equal = TRUE)
```
This p value is extremely small, so we can reject the null hypothesis that the means of the three groups are the same in favor of the alternative hypotheses that the means are not all the same.
</br>

# Problem 4 (Verzani problem 12.13)

*For the data in Problem 2, perform the one-way ANOVA using `lm()`. Compare to
the results of `oneway.test()`.*

```{r, error = TRUE}
# Your code here
```

# Problem 5 (Verzani problem 12.19)

*The `InsectSprays` data set contains a variable `count`, which counts the
number of insects and a factor `spray`, which indicates the treatment given.*

*First perform an analysis of variance to see whether the treatments make a
difference. If so, perform the Tukey HSD procedure to see which pairwise
treatments differ.*

```{r, error = TRUE}
# Your code here
```

