---
title: "MATH 3080 Lab Project 6"
author: "Your Name Here"
date: "12/27/2019"
output:
  pdf_document:
    toc: TRUE
---

# Problem 1 (Verzani problem 10.7)

*The more beer you drink, the more your blood alcohol level (BAL) rises. The
table below contains a data sete on beer consumption. Make a scatterplot with a
regression line and 95% prediction intervals drawn. Test the hypothesis that one
beer raises your BAL by 0.02% against the alternative that it raises it less.*

| beers |   5  |   2  |   9  |   8  |   3  |   7   |   3  |   5  |   3  |   5  |
|-------|------|------|------|------|------|-------|------|------|------|------|
| BAL   | 0.10 | 0.03 | 0.19 | 0.12 | 0.04 | 0.095 | 0.07 | 0.06 | 0.02 | 0.05 |

```{r, error = TRUE}
# Your code here
```

# Problem 2 (Verzani problem 10.11)

*For the `homedata` (**UsingR**) data set, find the regression equation to
predict the year-2000 value of a home from its year-1970 value. Make a
prediction for an \$80,000 home in 1970. Comment on the appropriateness of the
regression model by investigating the residuals.*

```{r, error = TRUE}
# Your code here
```

# Problem 3 (Verzani problem 10.13)

*The `alaska.pipeline` (**UsingR**) data set contains measurements of defects on
the Alaska pipeline that are taken first in the field and then in the
laboratory. The measurements are done in six batches. Fit a linear model for the
lab-defect size as modeled by the field-defect size. Find the coefficients.
Discuss the appropriateness of the model.*

```{r, error = TRUE}
# Your code here
```

# Problem 4 (Verzani problem 10.15)

*The `galton` (**UsingR**) data set contains data collected by Francis Galton in
1885 concerning the influence a parent's height has on a child's height. Fit a
linear model modeling a child's height by his parents'. Do a test of
significance to see whether $\beta_1$ equals 1 against a two-sided alternative.*

```{r, error = TRUE}
# Your code here
```

# Problem 5 (Verzani problem 10.17)

*The `alaska.pipeline` (**UsingR**) data set appears appropriate for a linear
model, but the assumption of equal variances does not seem appropriate. A
log-transformation of each variable does seem to have equal variances. Fit the
model*

$$\log(\texttt{lab.defect}) = \beta_0 + \beta_1 \log(\texttt{field.defect}) +
\epsilon.$$

*Investigate the residuals and determine whether the assumption of equal
variance seems appropriate.*

```{r, error = TRUE}
# Your code here
```

# Problem 6 (Verzani problem 10.19)

*In a simple linear regression, confidence intervals for $\beta_0$ and $\beta_1$
are given separately in terms of the $t$-distribution as $\hat{\beta}_i \pm t^*
\text{SE}(\hat{\beta}_i)$. They can also be found* jointly, *giving a
**confidence ellipse** for the parameters as a pair. This can be found easily in
R with the **ellipse** package. If `res` is the result of the `lm()` function,
then `plot(ellipse(res), type = "l")` will draw the confidence ellipse.*

*For the `deflection` (**UsingR**) data set, find the confidence ellipse for
`Deflection` modeled by `Load`.*

```{r, error = TRUE}
# Your code here
```
