---
title: "MATH 3080 Lab Project 7"
author: "Harrison Webb"
date: "02/27/2022"
output:
  html_document:
    toc: TRUE
    keep_md: TRUE
---

```{r}
#by julie sherman
predict_func <- function(fit) {
  response <- names(fit$model)[[1]]
  explanatory <- names(fit$model)[[2]]
  function(x, ...) {
    dat <- data.frame(x)
    names(dat) <- explanatory
    predict(fit, dat, ...)
  }
}

min_aic_model <- function(...) {
  models <- list(...)
  aic_list <- sapply(models, AIC)
  best_model <- models[[which.min(aic_list)]]
  best_model
}
```

# Problem 1 (Verzani problem 11.22)

*For the data set `trees`, model the `Volume` by the `Girth` and `Height`
variables. Does the model fit the data well?*

```{r, include=FALSE}
library(rgl)
library(UsingR)
library(MASS)
```


```{r, error = TRUE}
treeModel <- lm(Volume~Girth + Height, data = trees)
summary(treeModel)
```
3D Plot of the data for fun
```{r, fig.align='center', out.width='75%'}

plot3d(Volume~Girth + Height, data = trees)
rglwidget()
```

From the linear model, we see that the girth coefficient has an extremely small p-value and the height coefficient has a p value that is still significant at $\alpha < 0.05$ but is not nearly as small as the former. This indicates that both coefficients are useful. Additionally, we see that $\text{Adjusted }R^2 = 0.9442$, which indicates a good fit of the model to the data.
</br>

# Problem 2 (Verzani problem 11.24)

*For the `deflection` (**UsingR**) data set, fit the quadratic model*

$$\text{Deflection} = \beta_0 + \beta_1 \text{Load} + \beta_2 \text{Load}^2 +
\epsilon.$$

*How well does this model fit the data? Compare to the linear model.*

```{r, error = TRUE, fig.align='center', out.width='75%'}
data(deflection)

par(mfrow = c(1, 2))

deflectionModel <- lm(Deflection~Load + I(Load^2), data = deflection)
deflection_func <- predict_func(deflectionModel)
summary(deflectionModel)

deflectionModelLinear <- lm(Deflection~Load, data = deflection)

plot(Deflection~Load, data = deflection, main = "Quadratic Model")
lines(deflection$Load, deflection_func(deflection$Load))

plot(Deflection~Load, data = deflection, main = "Linear Model")
abline(deflectionModelLinear)

anova(deflectionModelLinear, deflectionModel)
```
Though both the linear and quadratic models give great fits, the data is clearly linear, so that model would be most appropriate. (in fact, the data is suspiciously linear, I almost feel like I have an error somewhere and am not plotting the right stuff??)

# Problem 3 (Verzani problem 11.25)

*The data set `kid.weights` contains age, weight, and height measurements for
several children. Fit the linear model*

$$\text{weight} = \beta_0 + \beta_1 \text{age} + \beta_2 \text{height} + \beta_3
\text{height}^2 + \beta_4 \text{height}^3 + \beta_5 \text{height}^4$$

*Use the partial $F$-test to select between this model and the nested models
found by using only first-, second-, and third-degree polynomials for `height`.*

```{r, error = TRUE}
weightsModel4 = lm(weight ~ age + height + I(height^2) + I(height^3) + I(height^4), data = kid.weights)
weightsModel3 = lm(weight ~ age + height + I(height^2) + I(height^3), data = kid.weights)
weightsModel2 = lm(weight ~ age + height + I(height^2), data = kid.weights)
weightsModel1 = lm(weight ~ age + height, data = kid.weights)

anova(weightsModel1, weightsModel2, weightsModel3, weightsModel4)
```
All F tests are statistically significant, so we can say that the full model is the best choice of model out of all 4.

# Problem 4 (Verzani problem 11.27)

*The data set `Cars93` (**MASS**) contains data on cars sold in the United
States in the year 1993. Fit a regression model with `MPG.city` modeled by the
numeric variables `EngineSize`, `Weight`, `Passengers`, and `Price`. Which
variables are marked as statistically significant by the marginal $t$-tests?
Which model is selected by the AIC?*

```{r, error = TRUE}
# Your code here
```

# Problem 5 (Verzani problem 11.29)

*The data set `baycheck` (**UsingR**) contains estimated populations for a
variety of Bay Checkerspot butterflies near California. A common model for
population dynamics is the Ricker model, for which $t$ is time in years:

$$N_{t + 1} = aN_t e^{bN_t} W_t,$$

*where $a$ and $b$ are parameters and $W_t$ is a lognormal multiplicative error.
This can be turned into a regression model by dividing by $N_t$ and then taking
logs of both sides to give*

$$\log\left(\frac{N_{t + 1}}{N_t}\right) = \log(a) + b N_t + \epsilon_t.$$

*Let $y_t$ be the left-hand side. This may be written as*

$$y_t = r\left(1 - \frac{N_t}{K}\right) + \epsilon_t,$$

*because $r$ can be interpreted as an unconstrained growth rate and $K$ as a
carrying capacity.*

*Fit the model to the `baycheck` data set and find values for $r$ and $K$. To
find $y_t$ you can do the following:*

```{r, error = TRUE}
d <- with(baycheck, {
  n <- length(year)
  yt <- log(Nt[-1]/Nt[-n])
  nt <- Nt[-n]
  data.frame(yt, nt)
})
```

*Recall that a negative index means all but that index.*

```{r, error = TRUE}
# Your code here
```

